Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-13 arxiv/setup.tex
typos

%Suppose we have $t$ labeled regression tasks. %where $t$ is a fixed value that does not grow with the feature dimension $p$.
%For each task $k$ from $1$ to $t$, we have $n_k$ samples $x_1, x_2, \dots, x_{n_k}$ that are all $p$-dimensional feature vectors with real-valued labels $y_1, y_2, \dots, y_{n_k}$.
%Without loss of generality, we focus on predicting the $t$-th task and refer the $t$-th task as the target.
%We refer to task $1$ to task $t-1$ as source tasks.
%We assume that the labels of each task $k$ satisfy the linear model with unknown parameters $\beta_k\in\real^p$, that is,
%	\[ y_i = x_i^{\top}\beta_k + e_i, \text{ for all } 1 \le i \le n_k, \]
%where $e_i$ denotes random noise with mean zero and variance $\sigma^2$.
%To write the above notations more succiently, let $X_k \in \real^{n_k \times p}$ denote the covariates that consists of a feature vector in every row.
%Let
%	\be\label{model_YvsX} Y^{(k)} = X^{(k)} \beta_k + \varepsilon_k, \text{ for all } 1\le k \le t \ee denote the label vector of task $k$ and $\varepsilon_k$ denote an i.i.d. random vector with mean zero and variance $\sigma^2$.
%Following \citet{HMRT19} and \citet{BLLT20},
%In the high-dimensional linear regression setting (e.g. ), the features of the $k$-th task, denoted by $X_k\in\real^{n_i\times p}$, consist of $n_k$ feature vectors given by $x_1, x_2, \dots, x_{n_k}$.
%we assume that for each task $k$ from $1$ to $t$, each feature vector $x_i = \Sigma^{1/2}_k z_i$ for $i = 1, \dots, n_k$, where $z_i\in\real^p$ is an i.i.d. random vector with  mean zero and unit variance.
%Recall that $\Sigma_k$ is the population covariance matrix task $k$'s feature vectors.
%Let $\Sigma_k = U_k D_k U_k^{\top}$ denote the singular value decomposition of $\Sigma_k$.
%We denote $\Sigma_k = U_k D_k^{1/2}$ as the square root of $\Sigma_k$.
%The sample size of task $k$, given by $n_k$, is equal to $\rho_k\cdot p$ for a fixed value $\rho_k$ that does not grow with feature dimension $p$.
%This setting is also known as the high-dimensional linear regression setting in the literature.
%There are two motivations for studying the high-dimensional linear regression setting for multi-task learning.
%First, this setting captures salient properties of modern large-scale datasets, where the sample sizes are usually on the order of tens to hundreds of the number of features \cite{sur2019modern}.
%Second, as we will see soon in Section \ref{sec_general}, we can derive precise asymptotics of the generalization error that scale with properties of the task data such as sample sizes.
%The labels $Y_k = X_k \beta_k + \varepsilon_k$, where $\beta_k$ denotes the linear model parameters and $\varepsilon_k$ denotes i.i.d. noise with mean zero and variance $\sigma^2$.
%Recall that we have $t$ labeled training datasets, denoted by $(X_1, Y_1), (X_2, Y_2), \dots, (X_t, Y_t)$, where $X_i\in\real^{n_i\times p}$ and $Y_i\in\real^{n_i}$ for $1\le i\le t$.
%Following \cite{HMRT19,BLLT20}, we assume that for each task $i = 1,2,\dots,t$,  every feature vector is generated as $x = \Sigma_i^{1/2} z$, where $z\in\real^p$ is a random vector with i.i.d. entries of mean zero and unit variance and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
%Without loss of generality, let the $t$-th task be the target task.

We assume that there are multiple datasets that all follow (potentially different) linear models.
In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.
We study a hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size, $X^{(i)} \in \real^{n_i \times p}$ denote its feature covariates, and $Y^{(i)} \in \real^{n_i}$ denote all the labels.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
Let $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
We study the following optimization objective.
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2.
\end{align}
Given a solution $(\hat{A}, \hat{B})$ to the above optimization problem, let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the hard parameter sharing (HPS) estimator for task $i$.
Some critical questions are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

%	\item Learn the output layers $W_1, W_2, \dots, W_t$: set $B = \hat{B}$ and minimize the training loss over $W_1, W_2, \dots, W_t$.
%		{\begin{align}\label{eq_mtl_eval}
%			g(W_1, \dots, W_t) = \sum_{k=1}^t \norm{X_k \hat{B} W_k - Y_k}^2.
%		\end{align}}
%\end{enumerate}
%Let $\hat{\beta}_t^{\MTL}$ denote the multi-task learning estimator obtained from the procedure above.
%By contrast, let $\hat{\beta}_t^{\STL} = (X_t^{\top}X_t)^{-1}X_t^{\top}{Y_t}$ denote the single-task learning estimator. % denoted by $L(\hat{\beta}_t^{\STL})$.
%where $B\in\real^{p\times r}$ and $W_k\in\real^r$ for every $1\le k\le t$.
%Following , we assume that $r < t$, because otherwise minimizing $f(\cdot)$ could result in $BW_i$ being the single-task optimum.
%\smallskip
%\noindent\textit{Remark.}
%In general, the multi-task learning objective $f(\cdot)$ is non-convex with respect to $B$ and $W_1, \dots, W_t$.
%Therefore, we first minimize $B$ in equation \eqref{eq_mtl} and then minimize $W_k$ given $B$ in equation \eqref{eq_mtl_eval}.
%For our results later in Section \ref{sec_general} and \ref{sec_special}, we will identify tractable cases and provide guarantees to the above procedure.
%\paragraph{Problem statement.}
%We focus on predicting a particular task, say the $t$-th task, without loss of generality.
%For an estimator $\hat{\beta}$ of the target task model $\beta_t$, we define the prediction loss of the estimator $\hat{\beta}$ as
%	{\begin{align}\label{eq1_prediction_loss}
%		\te(\hat{\beta}) = \exarg{x = \Sigma_t^{1/2} z}{({x}^{\top}\hat{\beta} - {x}^{\top}\beta_t)^2}
%		= \bignorm{\Sigma_t^{1/2} (\hat{\beta} - \beta_t)}^2,
%	\end{align}}%
%where $x = \Sigma_t^{1/2} z$ denotes a random feature vector with covariance $\Sigma_t$.
%In the above equation, $x^{\top}\beta_t$ is the true label of $x$.
%= (\hat{\beta} - \beta_t)^{\top}\Sigma_t(\hat{\beta} - \beta_t)
%We say that the source tasks provide a \textit{positive transfer} to the target task if the prediction loss of the MTL estimator is lower than that of the STL estimator, that is, if
%	\[ L(\hat{\beta}_t^{\MTL}) > L(\hat{\beta}_t^{\STL}). \]
%On the other hand, we say that the source tasks provide a \textit{negative transfer} to the target task if $L(\hat{\beta}_t^{\MTL}) < L(\hat{\beta}_t^{\STL})$.
%Our goal is to study when the source tasks provide a positive transfer to the target task.
%More specifically, we study how varying properties of task data including task similarity, sample ratio, and covariate shift affects information transfer in multi-task learning.

%We begin by defining our problem setup including the multi-task estimator we study.
%Then, we describe the bias-variance tradeoff of the multi-task estimator and connect the bias and variance of the estimator to \textit{task similarity}, \textit{sample size}, and \textit{covariate shift}.
%Finally, we show a tight concentration bound for the bias and variance quantities using random matrix theory.

%\subsection{Problem Formulation}

%In the worst case, the optimization objective \eqref{eq_mtl} is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
%Therefore, we work in cases where a global and/or local minimum characterization is possible

\paragraph{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (see e.g. \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp generalization bound on its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of hard parameter sharing admits a clean bias-variance decomposition.
Our results imply that hard parameter sharing helps by reducing variance compared to single-task learning, but hurts by increasing bias.
Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample size ratios and population covariance matrices.
We analyze the local minimizers of $f(A, B)$ and provide a sharp generalization bound on both tasks' prediction loss.
Despite its simplicity, we show several interesting phenomena by varying sample sizes and covariate shifts in this setting.
See Figure \ref{fig_intro_sample_size} for an illustration.

\begin{figure}[!t]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.5\textwidth,valign=t]{figures/mtl_model_arch.pdf}
		\caption{A hard parameter sharing architecture}
	\end{subfigure}\hfill
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.713\textwidth,valign=t]{figures/sample_ratio_c2_400.eps}
		\caption{Varying sample size ratio}
		\label{fig_intro_sample_size_b}
	\end{subfigure}
	\caption{An illustrative example of our result:
	Consider the prediction loss of hard parameter sharing (left) for task two, given two linear regression tasks.
	Increasing task one's sample size decreases task two's prediction loss initially, but increases afterwards. This phenomenon occurs due to different bias-variance tradeoffs as the sample size ratio increases. Our result provides an estimated loss (solid line) that accurately matches the empirical loss (dotted line).
	See Section \ref{sec_simulation} for the precise setting.}
	\label{fig_intro_sample_size}
\end{figure}

Consequently, using our precise loss estimates, we observe qualitative properties of hard parameter sharing for varying datasets' properties.
\begin{enumerate}
	\item \textit{Sample efficiency (Example \ref{ex_same_cov})}:
	One advantage of combining multiple datasets is that the requirement for labeled data reduces compared to STL, a phenomenon that \citet{ZSSGM18} has observed empirically.
	Our results further imply that HPS's sample efficiency depends on model-specific variance vs. noise variance. It is generally high when the noise variance is large compared to model-specific variance across different tasks.
	\item \textit{Sample size ratio (Example \ref{ex_sample_ratio})}: Increasing one task's sample size does not always help to reduce another task's loss. In a simplified setting, we find that the task loss either decreases first before increasing afterwards, or decreases monotonically depending on how fast the bias increases. These two trends result from different tradeoffs between increasing bias and decreasing variance.
	\item \textit{Covariate shift (Example \ref{ex_covshift})}: In addition to sample sizes, variance also scales with the covariate shift between different datasets. For a large sample size ratio, HPS's  variance is smallest when there is no covariate shift. Counterintuitively, for a small sample size ratio, having covariate shifts reduces variance through a complementary spectrum.
\end{enumerate}

@@

There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimum of $f(A, B)$ and its population version.
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. Using recently developed techniques from random matrix theory \cite{Anisotropic}, we show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}(i)). % Moreover, we can obtain a sharp bound on the concentration error.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}. 
%\HZ{add several sentences on the technical insight} 
One limitation of our analysis is that for two tasks with different sample sizes, there is an extra error term in the prediction loss of hard parameter sharing (cf. equation \eqref{cor_MTL_error}), which can be large for very small $n_1$. This requires studying the spectrum of non-symmetric random matrices, which is an intriguing open question (see Section \ref{sec_conclude} for more discussion).
% plus an identity matrix,  due to lack of freeness
%\HZ{add why this is challenging}

@@

There is a large body of both classical and recent works on multi-task learning.
We focus our discussion on theoretical works, and refer interested readers to several excellent surveys for general references \cite{PY09,R17,ZY17,V20}.
The early work of \citet{B00,BS03,M06} have sought for a study of multi-task learning from a theoretical perspective, often using uniform convergence or Rademacher complexity based techniques.
An influential paper by \citet{BBCK10} provides uniform convergence bound that combines multiple datasets in certain settings.
One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \citet{B00,MPR16}.
Moreover, these techniques do not apply to the high-dimensional setting, because the results usually require a sample size at least $p \log p$.

@@

\smallskip
\noindent\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we present our technical results that describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation} and \ref{sec_text}, we validate our theory in both simulations and a real world classification task.
In Section \ref{sec_conclude}, we conclude the paper and describe several open questions.

\paragraph{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For an $n\times p$ matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its spectral norm.
Let $\lambda_1(X) \ge \lambda_2(X) \ge \cdots \ge \lambda_{p\wedge n}(X)$ denote the singular values of $X$.
Let $X^+$ denote the Moore-Penrose psuedoinverse.
We refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.

>>>>

We assume that there are multiple datasets that all follow (potentially different) linear models.
In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.

We study a hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size, $X^{(i)} \in \real^{n_i \times p}$ denote its feature covariates, and $Y^{(i)} \in \real^{n_i}$ denote all the labels.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
Let $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
We study the following optimization objective.
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
\vspace{-0.1in}
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2.
\end{align}
%\vspace{-0.1in}
Given a solution $(\hat{A}, \hat{B})$ of $f(A, B)$, let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
Some critical questions are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\textbf{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (see e.g. \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp generalization bound on its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of hard parameter sharing admits a clean bias-variance decomposition.
Our results imply that hard parameter sharing helps by reducing the variance compared to single-task learning, but hurts by increasing the bias.
Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample size ratios and population covariance matrices.
We analyze the local minimizers of $f(A, B)$ and provide a sharp generalization bound on both tasks' prediction loss.
Despite its simplicity, we show several interesting phenomena by varying sample sizes and covariate shifts in this setting.
See Figure \ref{fig_intro_sample_size} for an illustration.

Consequently, using our precise loss estimates, we observe several qualitative properties of hard parameter sharing for varying datasets' properties.
%\begin{enumerate}

\textit{Sample efficiency (Example \ref{ex_same_cov})}:
	One advantage of combining multiple datasets is that the requirement for labeled data reduces compared to STL, a phenomenon that \citet{ZSSGM18} has observed empirically.
	Our results further imply that HPS's sample efficiency depends on model-specific variances across tasks vs. the noise variance, and is generally high when the latter is large.

\textit{Sample size ratio (Example \ref{ex_sample_ratio})}: Increasing one task's sample size does not always help to reduce another task's loss. In a simplified setting, we find that the task loss either decreases first before increasing afterwards, or decreases monotonically depending on how fast the bias increases. These two trends result from different bias-variance tradeoffs.

\textit{Covariate shift (Example \ref{ex_covshift})}: In addition to sample sizes, variance also scales with the covariate shift between different datasets. For a large sample size ratio, HPS's  variance is smallest when there is no covariate shift. Counterintuitively, for a small sample size ratio, having covariate shifts reduces variance through a complementary spectrum.
%\end{enumerate}

@@

There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimum of $f(A, B)$ and its population version.
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. Using recently developed techniques from random matrix theory \cite{Anisotropic}, we show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}). % Moreover, we can obtain a sharp bound on the concentration error.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}. 
%\HZ{add several sentences on the technical insight} 
One limitation of our analysis is that for two tasks with different sample sizes, there is an extra error term in the prediction loss of hard parameter sharing (cf. equation \eqref{cor_MTL_error}), which can be large for very small $n_1$. This requires studying the singular value spectrum of certain non-symmetric random matrices, which is an intriguing open question (see Section \ref{sec_conclude} for more discussion).
% plus an identity matrix,  due to lack of freeness
%\HZ{add why this is challenging}

@@

There is a large body of both classical and recent works on multi-task learning.
We focus our discussion on theoretical works, and refer interested readers to several excellent surveys for general references \cite{PY09,R17,ZY17,V20}.
The early work of \citet{B00,BS03,M06} have sought for a study of multi-task learning from a theoretical perspective, often using uniform convergence or Rademacher complexity based techniques.
An influential paper by \citet{BBCK10} provides uniform convergence bounds that combine multiple datasets in certain settings.
One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \citet{B00,MPR16}.
Moreover, these techniques do not apply to the high-dimensional setting, because the results usually require a sample size at least $p \log p$.

@@

\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we present our technical results that describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation} and \ref{sec_text}, we validate our theory in both simulations and a real world classification task.
In Section \ref{sec_conclude}, we conclude the paper and describe several open questions.

\textbf{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For an $n\times p$ matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its spectral norm.
Let $\lambda_1(X) \ge \lambda_2(X) \ge \cdots \ge \lambda_{p\wedge n}(X)$ denote the singular values of $X$.
Let $X^+$ denote the Moore-Penrose psuedoinverse.
We refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.
